Staged Abstract Interpreters

### 2019-09-04 paper/10_intro.tex
revision

\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
  \item We begin by introducing our target language and reviewing
    monads in Scala, and then presenting the generic interpreter
    (Section~\ref{prelim}), after which we review instantiations
    of concrete interpretation (Section~\ref{unstaged_conc}) and
    staged concrete interpretation (Section~\ref{stagedinterp}).
  \item We present the unstaged abstract interpreter under the same
    framework by replacing the environment, store, and values to their
    abstract counterparts (Section~\ref{unstaged_abs}). We then
    show that the combination of approximation and specialization, dubbed
    \textit{staged abstract interpreters}, can be readily derived
    (Section~\ref{sai}). We also summarize the approach and discuss
    soundness properties after showing the four artifacts.
  \item We conduct three case studies (Section~\ref{cases_study}) and
    an empirical evaluation on performance improvements (Section~\ref{evaluation}).
    Finally we discuss related work and conclude.
\end{itemize}

>>>>

\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
  \item We begin by introducing our target language and reviewing
    monads in Scala, and then presenting the generic interpreter
    (Section~\ref{prelim}), after which we review instantiations
    of concrete interpretation (Section~\ref{unstaged_conc}) and
    staged concrete interpretation (Section~\ref{stagedinterp}).
  \item We present the unstaged abstract interpreter under the same
    framework by replacing the environment, store, and values to their
    abstract counterparts (Section~\ref{unstaged_abs}). We then
    show that the combination of approximation and specialization, dubbed
    \textit{staged abstract interpreters}, can be readily derived
    (Section~\ref{sai}). We also summarize the approach and discuss
    soundness properties after showing the four artifacts.
  \item We summarize our approach and discuss correctness issues (Section~\ref{discussion}).
    After, we conduct three case studies (Section~\ref{cases_study}) and
    an empirical evaluation on performance improvements (Section~\ref{evaluation}).
    Finally we discuss related work and conclude.
\end{itemize}

### 2019-09-04 paper/40_absinterp.tex
revision

After seeing both unstaged and staged concrete interpreters, now we turn our
focus on abstract interpreters under the same framework. We first present a
lattice framework with binding-time abstraction, as well as simple abstract
domains such as power set lattice.
Then we show the abstract components, namely abstract values, abstract
environments, and abstract stores.
The abstract interpreter we construct in this
section is similar to \citet{DBLP:journals/pacmpl/DaraisLNH17}'s. For
simplicity, it is also context/path/flow-insensitive by our choice. In Section
\ref{cfa}, we will instantiate a commonly-used context-sensitive analysis;
\citet{Darais:2015:GTM:2814270.2814308} shows how to achieve path- and
flow-sensitivity by varying the monads, which is also applicable to our unstaged
or staged abstract interpreter.

\iffalse
For example, the CEK machine \cite{DBLP:conf/popl/FelleisenF87} for
concrete execution can be readily refactored to an effective $0$-CFA
control-flow analysis \cite{Shivers:1988:CFA:53990.54007,
  Midtgaard:2012:CAF:2187671.2187672}: first, tweak the environment
dereference as a nondeterministic choice, such that the environment
may contain multiple possible values for a variable, then allocate
continuations in the environment to exploit return-flow information,
and also constrain the address space to be finite.
\fi

@@

Abstract interpreters run on abstract domains, such abstract domains usually can
be presented by a lattice.
A complete lattice over set $S$ is a tuple of $\langle S, \sqsubseteq, \top,
\bot, \sqcup, \sqcap \rangle$, where $\top$ is the top element, $\bot$ is the
bottom element, $\sqsubseteq : S \to S \to Bool$ is the
ordering relation, $\sqcup: S \to S \to S$ is the join (least upper bound)
operator, and $\sqcap: S \to S \to S$ is the meet (greatest lower bound)
operator. The trait @SPLattice@ (Figure \ref{fig:splattice}) defines a type
class of stage polymorphic lattices. Similar to @MonadOps@ in Section
\ref{generic_if}, we introduce an additional higher-kinded type @R[_]@ to
annotate the binding-times, where @R[_]@ wraps datatype @S@ and @Boolean@ in
these operations.

@@

In this section, we simply let the binding-time type @R[T] = T@ (i.e., @NoRep@)
to instantiate an unstaged lattice type class. An example of lattice we used in
the rest of the paper is the power set abstract domain (shown in Figure
\ref{fig:powerset}), where the ordering is subset relation, the join is union
operation, and the meet is the intersect operation. The bottom element of a
power set is empty set, and we don't have a top representation for power set.
Other lattices used in the the paper, such as products and maps, can be
implemented similarly or lifted from existing lattices element-wise or
point-wise. Non-relational abstract domains such intervals also can be
implemented in a stage polymorphic way.

@@

We first widen the concrete values to @AbsValue@: numbers are lifted to a
singleton object @IntTop@ representing the set of all numbers; closures remain
the same. Then the type @Value@ is a set of @AbsValue@ (we abuse the type names
a bit, as the generic interface already defined the abstract type @Value@).
The address space is constrained to be finite to ensure the computability of
analysis -- for the simplest monovariant analysis, we use variable names for
addresses. After defining the @Value@ and @Addr@ type, the environments and
stores are automatically lifted to abstract environments and abstract stores,
i.e., @Map[Ident, Addr]@ and @Map[Addr, Set[AbsValue]]@ respectively.
It is important that the abstract store now maps addresses to sets of abstract
values, indicating that an address may point to an over-approximated set of
values occurred at run-time.
\begin{lstlisting}
  trait AbstractComponents extends Semantics {
    sealed trait AbsValue
    case object IntTop extends AbsValue
    case class CloV(lam: Lam, env: Env) extends AbsValue
    type Value = Set[AbsValue]; case class Addr(x: String)
    ...
  }
\end{lstlisting}

To prevent the abstract interpreter from falling into non-termination when
computing the fixed-point, the ADI approach uses a co-inductive mechanism
consisting of two caches that remember the input and output of functions.
Here we first provide the necessary type definitions, and later we will describe
the algorithm in detail. A @Cache@ is a mapping from configurations @Config@ to
sets of value-store pairs, where the configuration is a triple of current
expression being evaluated, environment and store. Intuitively, a cache memoizes
the corresponding values and stores for a given program configuration.
\begin{lstlisting}
  type Config = (Expr, Env, Store); type Cache = Map[Config, Set[(Value, Store)]]
\end{lstlisting}

Now, the store is a mapping from addresses to sets of abstract values. This can
be justified by the approximation nature of abstract interpretation and
nondeterminism in the abstract interpreter.
First, when analyzing a conditional expression, we may not have enough
information on which branch will be taken, thus a sound treatment is to
nondeterministically explore both branches, i.e., compute the join of two path.
Second, when applying a function, for instance, @f(a)@, the possible lambda
terms of @f@ is computed approximately. As a result, we may retrieve a set of
closures when dereferencing the address of @f@ and need to apply all of them.

\paragraph{Monads for Abstract Interpretation} Compared with the concrete
interpreter that uses reader and state effects, the abstract interpreter
further introduces a nondeterminism effect and another reader and state effect
for the two caches. The nondeterminism effect is represented by the
@Set[M[_], A]@ monad transformer, where @M@ is the inner monad being
transformed, and @A@ is the type of elements in the set. We use a @ReaderT@ for
one cache that will not be changed during one fixed-point iteration, and use a
@StateT@ for another cache that will be constantly updated during analysis.
\citet{DBLP:journals/pacmpl/DaraisLNH17} discuss different permutation of the
monad stack for abstract interpretation. The following @AnsM@ type shows the
monad stack (we use \hl{light gray} to highlight what have been changed from
the monad stack of concrete interpretation):
\begin{lstlisting}[escapechar=!]
trait AbstractSemantics extends AbstractComponents {
  type R[T] = T
  type AnsM[T] = ReaderT[StateT[!\hl{SetT}![!\hl{ReaderT}![!\hl{StateT}![IdM, !\hl{Cache}!, ?], !\hl{Cache}!, ?], ?], Store, ?], Env, T]
  ...
}
\end{lstlisting}

@@

Such monad stack scheme leads to a different type of result type. In the concrete
setting, we only have a reader monad (for environment) and a state monad (for
store), and the reader effect is not persistent to the final result. Together
with the value produced by the interpreter, the final result type is a pair of
@Value@ and @Store@. In the abstract setting, we have a @SetT@ inside of the
environment monad and store monad, which becomes the container type of the pair
of values and stores, i.e., @Set[(Value, Store)]@. Also note that the reader and
state monad for caches live internally of the nondeterminism monad, therefore
the cache type of state monad becomes adjacent to the set type.
\begin{lstlisting}
  type Result = (R[Set[(Value, Store)]], R[Cache])
\end{lstlisting}

>>>>

After seeing the unstaged and staged concrete interpreter, now we turn our
focus on abstract interpreters under the same framework. We first present a
lattice representation with binding-time abstraction, as well as simple
abstract domains such as power set lattice.
Then we show the abstract components, namely abstract values, abstract
environments, and abstract stores.
The abstract interpreter we construct in this
section is similar to \citet{DBLP:journals/pacmpl/DaraisLNH17}'s. For
simplicity, it is also context/path/flow-insensitive by our choice. In Section
\ref{cfa}, we will instantiate it to a commonly-used context-sensitive analysis;
\citet{Darais:2015:GTM:2814270.2814308} shows how to achieve path- and
flow-sensitivity by varying the monads, which is also applicable to our unstaged
or staged abstract interpreter.

@@

Abstract interpreters operate on abstract domains, such abstract domains
usually can be formulated by complete lattices.
A complete lattice over set $S$ is a tuple of $\langle S, \sqsubseteq, \top,
\bot, \sqcup, \sqcap \rangle$, where $\top$ is the top element, $\bot$ is the
bottom element, $\sqsubseteq : S \to S \to Bool$ is the
ordering relation, $\sqcup: S \to S \to S$ is the join (least upper bound)
operator, and $\sqcap: S \to S \to S$ is the meet (greatest lower bound)
operator. The trait @SPLattice@ (Figure \ref{fig:splattice}) defines a type
class of stage polymorphic lattices. Similar to @MonadOps@ in Section
\ref{generic_if}, we introduce an additional higher-kinded type @R[_]@ to
annotate the binding-times, where @R[_]@ wraps datatype @S@ and @Boolean@ in
these operations.

@@

In this section, we simply let the binding-time type @R[T] = T@ (i.e., @NoRep@)
to instantiate an unstaged lattice type class. An example of lattice we used in
the rest of the paper is the power set abstract domain (shown in Figure
\ref{fig:powerset}): the ordering is subset relation, the join is the union
operation, and the meet is the intersect operation. The bottom element of a
power set is empty set, and we don't have a top representation for power set.
Other lattices used in the paper, such as products and maps, can be
implemented similarly or lifted from existing lattices element-wise or
point-wise. Non-relational abstract domains such intervals can also be
implemented in a stage polymorphic way.

@@

We first widen the concrete values to @AbsValue@: numbers are lifted to a
singleton object @IntTop@ representing the set of all numbers; closures remain
the same. Then the type @Value@ is a set of @AbsValue@ (we abuse the type names
a bit, as the generic interface already defined the abstract type @Value@).
The address space is constrained to be finite to ensure the computability of
analysis -- for the simplest monovariant analysis, we use variable names for
addresses. After defining the @Value@ and @Addr@ type, the environments and
stores are automatically lifted to abstract environments and abstract stores,
i.e., @Map[Ident, Addr]@ and @Map[Addr, Set[AbsValue]]@, respectively.
\begin{lstlisting}
  trait AbstractComponents extends Semantics {
    sealed trait AbsValue
    case object IntTop extends AbsValue
    case class CloV(lam: Lam, env: Env) extends AbsValue
    type Value = Set[AbsValue]; case class Addr(x: String)
    ...
  }
\end{lstlisting}

It is important that the abstract store now maps addresses to sets of abstract
values, indicating that an address may point to an over-approximated set of
values occurred at run-time.  This can be justified by the approximation nature
of abstract interpretation and nondeterminism in the abstract interpreter.
First, when analyzing a conditional expression, we may not have enough
information on which branch will be taken, thus a sound treatment is to
nondeterministically explore both branches, i.e., compute the join of two path.
Second, when applying a function, for instance, @f(a)@, the possible lambda
terms of @f@ is computed approximately. As a result, we may retrieve a set of
closures when dereferencing the address of @f@ and need to apply all of them.

To prevent the abstract interpreter from falling into non-termination when
computing the fixed-point, the ADI approach uses a co-inductive mechanism
consisting of two caches that remember the input and output of functions.
Here, we first provide the necessary type definitions, and later we will describe
the algorithm in detail. A @Cache@ is a mapping from configurations @Config@ to
sets of value-store pairs, where the configuration is a triple of current
expression being evaluated, environment and store. Intuitively, a cache memoizes
the corresponding values and stores for a given program configuration.
\begin{lstlisting}
  type Config = (Expr, Env, Store); type Cache = Map[Config, Set[(Value, Store)]]
\end{lstlisting}

\paragraph{Monads for Abstract Interpretation} Compared with the concrete
interpreter that uses reader and state effects, the abstract interpreter
further introduces a nondeterminism effect and another reader and state effect
for the two caches. The nondeterminism effect is represented by the
@Set[M[_], A]@ monad transformer, where @M@ is the inner monad being
transformed, and @A@ is the type of elements in the set. We use a @ReaderT@ for
one cache that will not be changed during one fixed-point iteration, and use a
@StateT@ for another cache that will be constantly updated during the analysis.
\citet{DBLP:journals/pacmpl/DaraisLNH17} discuss different permutation of the
monad stack for abstract interpretation. The following @AnsM@ type shows the
monad stack (we use \hl{light gray} to highlight what have been changed from
the monad stack of concrete interpretation):
\begin{lstlisting}[escapechar=!]
trait AbstractSemantics extends AbstractComponents {
  type R[T] = T
  type AnsM[T] = ReaderT[StateT[!\hl{SetT}![!\hl{ReaderT}![!\hl{StateT}![IdM, !\hl{Cache}!, ?], !\hl{Cache}!, ?], ?], Store, ?], Env, T]
  ...
}
\end{lstlisting}

@@

Such monad stack scheme leads to a different type of grounded result. In the concrete
setting, we only have a reader monad (for environment) and a state monad (for
store), and the reader effect is not persistent to the final result. Together
with the value produced by the interpreter, the final result type is a pair of
@Value@ and @Store@. In the abstract setting, we have a @SetT@ inside of the
environment monad and store monad, which becomes the container type of pairs
of values and stores, i.e., @Set[(Value, Store)]@. Also note that the reader and
state monad for caches inhibit internally of the nondeterminism monad, therefore
the cache type of state monad becomes adjacent to the set type.
\begin{lstlisting}
  type Result = (R[Set[(Value, Store)]], R[Cache])
\end{lstlisting}
